The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-03-19 paper/iphas-dr2.tex
Adds citations to introduction + minor fixes

\author[G. Barentsen
et. al]{G. Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, 
R. Greimel$^3$, \newauthor
M. J. Irwin$^2$,
B. Mizalski$^{4}$,
C. Ruhland$^1$,
P. Groot$^5$,
A. Mampaso$^6$, 
S. E. Sale$^7$, \newauthor
M. J. Barlow$^{8}$,
R. L. M Corradi$^{6}$,
J. J. Drake$^{9}$,
J. Fabregat$^{10}$, 
B. T. Gaensicke$^{11}$, \newauthor
A. S. Hales$^{12}$,
J. Irwin$^{9}$,
C. Knigge$^{13}$,
T. Kupfer$^{5}$,
D. J. Lennon$^{14}$, 
J. R. Lewis$^{2}$, \newauthor
M. Mohr-Smith$^{1}$, 
R. A. H. Morris$^{15}$,
T. Naylor$^{16}$,
Q. A. Parker$^{17}$, 
S. Phillipps$^{13}$, \newauthor
R. Raddi$^{11}$, 
P. Rodriguez-Gil$^{6}$,
L. Sabin$^{18}$,
S. Scaringi$^{19}$,
D. Steeghs$^{11}$, 
Y. C. Unruh$^{20}$, \newauthor
K. Viironen$^{21}$, 
J. S. Vink$^{22}$, 
N. A. Walton$^{2}$,
N. J. Wright$^{1}$,
A. A. Zijlstra$^{23}$.
\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$South African Astronomical Observatory, P.O. Box 9, Observatory, 7935 Cape Town, South Africa.\\
$^{5}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{6}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{7}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{8}$University College London, Department of Physics \& Astronomy, 
Gower Street, London WC1E 6BT, U.K.\\
$^{9}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 
Cambridge, MA 02138, U.S.A. \\
$^{10}$Observatorio Astr\'onomico, Universidad de Valencia,
Catedr\'atico Jos\'e Beltr\'an 2, 46980 Paterna, Spain\\
$^{11}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{12}$Joint ALMA Observatory, Alonso de Córdova 3107, Vitacura 763-0355, Santiago, Chile. \\
$^{13}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
$^{14}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
$^{15}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
$^{16}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
$^{17}$Department of Physics \& Astronomy, Macquarie University, NSW 2109, Australia\\
$^{18}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
$^{19}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium\\
$^{20}$Department of Physics, Blackett Laboratory, Imperial College
London, Prince Consort Road, London, SW7 2AZ, U.K.\\
$^{21}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
$^{22}$Armagh Observatory, College Hill, Armagh, Northern Ireland,
BT61 9DG, U.K.\\
$^{23}$Jodrell Bank Centre for Astrophysics, School of Physics \&
Astronomy, University of Manchester, Manchester M13 9PL, U.K.}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against SDSS,
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ colour-colour diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure 
of our own backyard, the Milky Way.
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}, 
to be extended into larger, deeper, more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission line objects \citep{Witham2008};
discoveries of new northern symbiotic stars \citep{Corradi2008, Corradi2010}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects \citep{Vink2008,Barentsen2011a};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Corradi2011, Viironen2011}.

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7 arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7 arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$/\ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\%of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy 
ensures that every pointing is immediately followed 
by an offset pointing at a displacement of $+$5 arcmin in Dec 
and $+$5 arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were determined empirically by inspecting
the images and photometry by eye.
We rejected 11\% of the data as part of this step
(many of which had already been rejected as part of the limiting magnitude criterion).

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
%(e.g. bright satellite trails)
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1 arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands.
In the majority of the cases we were able to correct the astrometry
of such fields by hand.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field (H$\alpha$/$r$/$i$) 
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at the $\sim$same time --  
essentially always within an elapsed time of 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap did not exceed 48 hours.
We note that the exact epoch of the magnitude in each band
is included in the source catalogue
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[2MASS;][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[SDSS;][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
and Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

@@

In future work, we intend to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We also aim to correct the IPHAS photometry for the
radial field distortions,
plan to produce a catalogue 
which details all the secondary detections
to aid time-domain studies,
and are investigating the use of photometry
derived from PSF-fitting across the survey.
%TODO: mention Gaia, e.g. IPHAS for transient identification?

@@

%IPHAS also provide images for some of the most
%crowded regions to be faced by Gaia.

>>>>

\author[G. Barentsen
et. al]{G. Barentsen$^{1}$\thanks{E-mail:geert@barentsen.be},
H. J. Farnhill$^1$,
J. E. Drew$^1$,
E. A. Gonz$\acute{\rm{a}}$lez-Solares$^2$, 
R. Greimel$^3$, \newauthor
M. J. Irwin$^2$,
B. Mizalski$^{4}$,
C. Ruhland$^1$,
P. Groot$^5$,
A. Mampaso$^6$, 
S. E. Sale$^7$, \newauthor
M. J. Barlow$^{8}$,
R. L. M Corradi$^{6}$,
J. J. Drake$^{9}$,
J. Eisl\"offel$^{10}$,
J. Fabregat$^{11}$, \newauthor
B. T. Gaensicke$^{12}$,
A. S. Hales$^{13}$,
J. Irwin$^{9}$,
C. Knigge$^{14}$,
T. Kupfer$^{5}$,
D. J. Lennon$^{15}$, \newauthor
J. R. Lewis$^{2}$, 
M. Mohr-Smith$^{1}$, 
R. A. H. Morris$^{16}$,
T. Naylor$^{17}$,
Q. A. Parker$^{18}$, \newauthor
S. Phillipps$^{13}$, 
R. Raddi$^{11}$, 
P. Rodriguez-Gil$^{6}$,
L. Sabin$^{19}$,
S. Scaringi$^{20}$,
D. Steeghs$^{11}$, \newauthor
Y. C. Unruh$^{21}$, 
K. Viironen$^{22}$, 
J. S. Vink$^{23}$, 
N. A. Walton$^{2}$,
N. J. Wright$^{1}$,
A. A. Zijlstra$^{24}$.
\\
$^{1}$School of Physics, Astronomy \& Mathematics, University of Hertfordshire, College Lane, Hatfield, Hertfordshire, AL10 9AB, U.K.\\
$^{2}$Institute of Astronomy, Cambridge University, Madingley Road, Cambridge, CB3 OHA, U.K.\\
$^{3}$IGAM, Institute of Physics, University of Graz, Universit\"atsplatz 5, Graz, Austria.\\
$^{4}$South African Astronomical Observatory, P.O. Box 9, Observatory, 7935 Cape Town, South Africa.\\
$^{5}$Afdeling Sterrenkunde, Radboud Universiteit Nijmegen, Faculteit NWI, Postbus 9010, 6500 GL Nijmegen, The Netherlands.\\
$^{6}$Instituto de Astrof\'isica de Canarias, 38200 La Laguna, Tenerife, Spain.\\
$^{7}$Rudolf Peierls Centre for Theoretical Physics, Keble Road, Oxford, OX1 3NP, U.K.\\
$^{8}$University College London, Department of Physics \& Astronomy, 
Gower Street, London WC1E 6BT, U.K.\\
$^{9}$Harvard-Smithsonian Center for Astrophysics, 60 Garden Street, 
Cambridge, MA 02138, U.S.A. \\
$^{10}$Th\"uringer Landessternwarte, Sternwarte 5, 07778, Tautenburg, Germany \\
$^{11}$Observatorio Astr\'onomico, Universidad de Valencia,
Catedr\'atico Jos\'e Beltr\'an 2, 46980 Paterna, Spain\\
$^{12}$Department of Physics, University of Warwick, Gibbet Hill Road, Coventry, CV4 7AL, U.K.\\
$^{13}$Joint ALMA Observatory, Alonso de Córdova 3107, Vitacura 763-0355, Santiago, Chile. \\
$^{14}$School of Physics \& Astronomy, University of Southampton, Southampton, SO17 1BJ, U.K.\\
$^{15}$European Space Astronomy Centre (ESAC), Villafranca del Castillo, Villanueva de la Canada, E-28692 Madrid, Spain.\\
$^{16}$School of Physics, Bristol University, Tyndall Avenue, Bristol, BS8 1TL, U.K.\\
$^{17}$School of Physics, University of Exeter, Stocker Road, Exeter, EX4 4QL, U.K.\\
$^{18}$Department of Physics \& Astronomy, Macquarie University, NSW 2109, Australia\\
$^{19}$Instituto de Astonom\'ia y Meteorolog\'ia, Departamento de F\'isica, CUCEI, Universidad de Guadalajara, Mexico.\\
$^{20}$Instituut voor Sterrenkunde, K.U. Leuven, Celestijnenlaan 200D, B-3001 Leuven, Belgium\\
$^{21}$Department of Physics, Blackett Laboratory, Imperial College
London, Prince Consort Road, London, SW7 2AZ, U.K.\\
$^{22}$Centro de Estudios de F\'isica del Cosmos de Arag\'on, Plaza San Juan 1, Planta 2, Teruel, 44001, Spain. \\
$^{23}$Armagh Observatory, College Hill, Armagh, Northern Ireland,
BT61 9DG, U.K.\\
$^{24}$Jodrell Bank Centre for Astrophysics, School of Physics \&
Astronomy, University of Manchester, Manchester M13 9PL, U.K.}

@@

\begin{abstract} % BACKGROUND, OBJECTIVE, METHODS, RESULTS, CONCLUSIONS
The INT/WFC Photometric H$\alpha$ Survey 
of the Northern Galactic Plane (IPHAS)
is a 1800 deg$^2$ imaging survey
covering the entire northern Milky Way at $|b| < 5^\circ$
in the $r$, $i$ and \ha\ filters 
using the Wide Field Camera (WFC) 
on the 2.5-metre Isaac Newton Telescope (INT) in La Palma.
We present the first quality-controlled and
uniformly-calibrated source catalogue
to have been extracted from the images,
providing single-epoch photometry
for 219 million unique sources
across 92\% of the survey area.
The observations were carried out between 2003 and 2013
at a median seeing of 1.1 arcsec
and to a mean $5\sigma$-depth of 
21.2 ($r$), 20.0 ($i$) and 20.3 ($\ha$)
in the Vega magnitude system.
We explain the data reduction 
and quality control procedures,
describe and test the new uniform calibration,
and detail the construction of the new 97-column source catalogue.
We show that the global calibration is accurate to
0.03~mag (rms) by validating against SDSS,
and provide a series of recommended quality criteria
which can be used to select the most reliable data
from the catalogue.
Finally, we demonstrate the ability of the 
catalogue's unique
$(r-\ha,\ r-i)$ diagram to
(i) characterise stellar populations and extinction regimes
towards different Galactic sight-lines
and (ii) select candidate \ha\ emission-line objects.
IPHAS is the first survey to offer comprehensive CCD photometry
of point sources across the Galactic Plane at visible wavelengths,
providing the much-needed counterpart
to recent infrared surveys.
\end{abstract}

@@

\section{Introduction}
The INT/WFC Photometric H$\alpha$ Survey
of the Northern Galactic Plane \citep[IPHAS;][]{Drew2005}
is providing new insights into the contents and structure 
of our own backyard, the Milky Way.
The original motivation for undertaking 
this large-scale programme of observation
-- spanning almost a decade 
and using more than 300 nights 
at the Isaac Newton Telescope (INT) in La Palma -- 
was to provide the digital update 
to the photographic northern H$\alpha$ surveys 
of the mid-20th century. 
By increasing the sensitivity 
with respect to these previous surveys 
by a factor $\sim$1000 (7 magnitudes), 
it was envisaged that IPHAS would allow 
the limited bright samples of Galactic emission line objects 
available at the outset \citep[e.g.][]{Kohoutek1999}, 
to be extended into larger, deeper, more statistically-robust samples.
This in turn could better inform our understanding 
of the early and late stages of stellar evolution.
Since the publication of the IPHAS Initial Data Release \citep[IDR;][]{Gonzalez-Solares2008},
these aims have begun to be realised through a
range of published studies including: 
a preliminary catalogue of candidate emission line objects \citep{Witham2008};
discoveries of new symbiotic stars \citep{Corradi2008, Corradi2010, Corradi2011}; 
new cataclysmic variables \citep{Witham2007}; 
new groups of young stellar objects
\citep{Vink2008,Barentsen2011a,Wright2012};
%new asymptotic giant branch stars \citep{Wright2008};
new classical Be stars \citep{Raddi2013};
along with discoveries of new and remarkable planetary nebulae 
\citep{Mampaso2006, Corradi2011, Viironen2011}
and new supernova remnants \citep{Sabin2013}.

@@

(3) \emph{Seeing.} 
The survey originally aimed to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at early longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Figure~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$/\ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec
and 10\% is better than 0.8~arcsec.
To improve the area covered by this data release,
we have decided to include the small fraction of data
that was obtained under seeing up to 2.5 arcsec,
so that only 1\%of the pipelined exposures
had to be excluded due to poor seeing.
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

@@

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy 
ensures that every pointing is immediately followed 
by an offset pointing at a displacement of $+$5~arcmin in Dec 
and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were determined empirically by inspecting
the images and photometry by eye.
We rejected 11\% of the data as part of this step
(many of which had already been rejected as part of the limiting magnitude criterion).

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
%(e.g. bright satellite trails)
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1~arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands.
In the majority of the cases we were able to correct the astrometry
of such fields by hand.

@@

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed at the $\sim$same time --  
essentially always within an elapsed time of 5 minutes.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

@@

Despite the best efforts made to obtain a nightly calibration,
large surveys naturally possess field-to-field variations
at the level of 0.1 mag
due to atmospheric changes during the night
and imperfections in the pipeline or the instrument
(e.g. the WFC is known to suffer from sporadic errors
in the timing of exposures).
Such variations need to be corrected for 
during a global re-calibration procedure.
Notable past examples include the global re-calibration 
of the Two Micron All Sky Survey \citep[2MASS;][]{Nikolaev2000},
the Sloan Digital Sky Survey \citep[SDSS;][]{Padmanabhan2008}
and the Panoramic Survey Telescope 
And Rapid Response System survey \citep[Pan-STARRS;][]{Schlafly2012},
which all achieved photometry 
that is globally consistent to within 0.01--0.02~mag.

@@

In future work, we intend to draw upon the PanSTARRS photometric
survey to further improve the accuracy of our global calibration.
We also aim to correct the IPHAS photometry for the
radial field distortions,
plan to produce a catalogue 
which details all the secondary detections
to aid time-domain studies,
and are investigating the use of photometry
derived from PSF-fitting across the survey.

@@
